A Guide to Designing Experiments to Test Statistical Graphics

### 2025-03-26 index.tex
Update index and diffs

\title{A Guide to Designing Experiments to Test Statistical Graphics}
\author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
\date{2025-03-16}

@@

While there are heuristics, rules-of-thumb, and guidelines (Allen \&
Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on Standards
for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating useful
and visually attractive data displays, the best way to establish the
efficacy of various design decisions is to test the visualization on
humans, evaluating different variants under controlled conditions
(Cleveland et al., 1988; Cleveland \& McGill, 1985). Empirical
assessments of visualizations, when carefully designed, allow
statisticians to determine which representation of the same data is most
effective along one or more dimension(s) of interest: estimation or
prediction accuracy, within or between group comparisons, response time,
and ability to draw real-world decisions are common goals for charts.

@@

There are many different testing methods used to empirically assess
statistical graphics. This paper uses studies conducted online without
additional equipment as primary examples, though many of the same
considerations apply to in-person experiments conducted using additional
equipment, including 3D printed charts, eye-tracking equipment, and
interactive data displays. Online experiments have lower overhead, offer
relatively fast data collection, and provide useful results for
well-designed experiments. The toolkit used for these experiments is
R-based (R Core Team, 2022), and includes ggplot2 (Wickham, 2016) and
Shiny (Chang et al., 2021) as primary components. In many experiments,
we customized the Shiny interface with JavaScript and d3 (Bostock et
al., 2011), enabling interactive graphics, use of svgs, and other useful
extensions. While we prefer this set of tools, most of the observations
described here apply to a wide variety of different workflows for
graphical experimentation, including in-person experiments.

@@

Visual inference (Buja et al., 2009; Wickham et al., 2010) is another
useful testing tool for perceptual questions such as ``which chart
displays this data more clearly'' (Hofmann et al., 2012) while
simultaneously assessing the statistical significance of the graphical
finding in a chart. Visual inference charts are often called ``lineups''
in analogy to the criminal procedure where the suspect is placed in a
line with several other individuals with similar characteristics. In a
graphical lineup procedure, there is a target plot containing the real
data, embedded in an array of innocent ``null'' plots generated through
resampling or simulation. If viewers consistently pick the target plot
at a higher rate than any of the null plots, the target plot is said to
be visually significant (Loy \& Hofmann, 2013; Majumder et al., 2013)
and a ``see'' value, the visual analogue of a \(p\)-value (Chowdhury et
al., 2020), can be calculated using the \texttt{vinference} R package or
the process described in Vanderplas et al. (2021). The details of this
calculation are beyond the scope of this broader discussion of how to
test charts, but more detail on visual inference is provided in
\textless insert citation to visual inference WIRE article under
development\textgreater.

@@

Another useful measurement strategy is to require participants to
\emph{engage directly} with an interactive visualization. This is useful
in a directed task, where users are asked to interact with the chart in
a specific way and the result is recorded, but it is also possible to
use interactive visualizations in an open-ended task, recording how
users engage with the graphic in an exploratory (as opposed to
goal-directed) manner. In one recent experiment, we asked participants
to forecast an exponential trend, with data presented on either a linear
or log scale. Using JavaScript code modified from New York Times
interactive graphics ``You Draw It'' features (Katz, 2017), we had users
draw trend lines with their computer mouse and make forecasts directly
on interactive charts, with the data and user-drawn predictions recorded
to our database. With interactive graphics rendered using JavaScript (or
other web libraries), the only limit to the types of questions one can
ask in testing graphics is one's ability to write code to interact with
the visualization library. This type of testing method can be extremely
natural for participants, but it also is hard to generalize when
discussing testing methods because of the potential range of
applications where it might be employed.

@@

All of the testing methods described here require significant work to
develop a strategy for data generation appropriate for testing the
underlying question. For instance, when testing the perception of
exponential growth, we had to develop a model which would generate data
with varying growth rates, but where the data had a pre-specified domain
and range. The data generating model is particularly critical when using
lineups, as the null sampling model must replicate the important visual
features in the data. Each testing method has specific requirements, but
it is important to carefully calibrate the model parameters to allow for
some variability, but not too much, and to ensure that participants can
succeed at the task and do not feel like they are being made to analyze
random noise. This Goldilocks-style problem is the focus of the next
section.

@@

The choice of the parameter space used in testing is crucial to gain
insight from a study without putting too much burden on participants
with overlong studies. Choosing an appropriate space for testing
parameters is a well-known problem in psychometric testing: the space
considered should cover the area between `only some activation' to
`almost full activation' of an appropriate psychometric function (Schütt
et al., 2016; Valentin et al., 2024). When testing charts, visual
assessment is obviously key, but researchers can make use of statistical
indices related to the testing condition to narrow the parameter space
to a reasonable and efficient subset from which maximal information can
be acquired.\\
These statistical indices may also serve as quantitative proxies for the
difficulty of the visual task. To identify a statistical proxy for
visual difficulty that may help with narrowing the parameter space, it
can be useful to consider numerical measures used to estimate the same
types of visual information that will be assessed in the experiment. For
instance, we have used

\begin{itemize}
\tightlist
\item
  \(R^2\) as a measure of the strength of a linear relationship
\item
  Gini inequality as a measure of the strength of clustering
\item
  lack-of-fit statistics to assess the amount of curvature in an
  exponential relationship (shown in
  Figure~\ref{fig-lof-density-curves})
\end{itemize}

@@

We cannot overstate the importance of visual assessment of your model
stimuli, preferably with fresh eyes. We highly recommend performing
several rounds of think-aloud pilot testing (focus) before deploying an
experiment. In support of this assessment, we offer up a cautionary tale
of our own experience: that of Vanderplas \& Hofmann (2017), where we
designed an experiment to test which plot aesthetics promoted discovery
of linear trends and/or clusters.

@@

Rice et al. (2024) recruited participants using nationally
representative panel samples and found that conclusions from fully
representative samples of the population can be very different from
volunteer samples recruited using other services, as many people are
unmotivated to engage with charts, don't know how to read charts, or
impose pre-formed conclusions onto visual displays. While using
representative panel sampling products offered by organizations such as
NORC or Gallup is more expensive than samples from Amazon MTurk or
Prolific, these results may imply that results from less representative
participant recruitment methods may not generalize to the broader
population. People who participate in studies via online platforms are
more technologically sophisticated and educated than the general
population, and this bias may significantly impact the conclusions.
However, it is also reasonable to argue that higher education might make
someone more likely to use charts and data for decision-making.\\
Researchers designing a study should carefully consider whether the goal
of the study is to generalize results to the adult population or to a
subset of that population who make decisions based on data.

@@

One highly useful (but not strictly essential) component of an
experiment that can be set up during the pilot testing stage is a basic
analysis script which summarizes all data collected to date visually. We
have used such scripts in the past to produce automatically updating
dashboards or web pages, allowing for real time or near-real time
monitoring of data collection efforts. This provides an easy way to
summarize completion of the experiment so that individuals can receive
credit (if using services like MTurk or Prolific), but also allows
interested participants to see individual results, which can be a factor
when recruiting on social media. As data collection online can also
happen extremely quickly (300 participants in \textless2h in our most
recent Prolific experiment), this can improve data monitoring, allowing
any issues to be spotted and resolved quickly. If server load is a
potential issue, it may also help to release batches of trials over a
longer period of time in order to minimize the chance of having to make
participants wait for others to complete the task before the server can
handle additional connections\footnote{This is the one major drawback to
  our preferred solution of self-hosting a Shiny server to handle data
  collection: the free version of Shiny server is limited to about 15
  connections at any given time. Prolific has recently added
  rate-limiting functions to the experimental control platform, which
  makes controlling the number of active jobs much easier.}. Batch trial
releases can also be used to ensure that participants are recruited
across different time zones; in some studies this is beneficial, while
in others it may be more important to control the release time to target
individuals in e.g.~North America rather than Europe.

@@

The see-value approach (Chowdhury et al., 2020; Majumder et al., 2013;
Vanderplas et al., 2021) is extremely useful for single lineups, but
when a series of lineups that are part of a designed experiment are
used, generalized linear mixed models are a much simpler way to
summarize the overall effect of various manipulated factors (Hofmann et
al., 2012; Robinson, 2022; Vanderplas \& Hofmann, 2017). This approach
also works extremely well for psychometric experiments (Vanderplas \&
Hofmann, 2015), as psychometric models can be easily fit into the
framework of a generalized linear mixed model (Ju et al., 2024) that has
higher power than the Rasch models (Andrich, 1988; Lu et al., 2022)
which were historically recommended for these experiments. In addition,
similar model structures with different link functions can be used to
model accuracy, response time, and confidence, if all three types of
information are collected from participants during the experiment.

@@

While Figure~\ref{fig-density-rug} shows point estimates, the same
approach can be modified to account for experimental methods which
generate participant response curves. In these cases, rather than adding
a rug plot to a density plot showing aggregate estimates, it may be more
helpful to create spaghetti plots of individual estimates with a
superimposed consensus estimate and relevant annotations showing
e.g.~anchor points.\\
The advantage of this approach is that it can accommodate extremely
messy data without requiring the extensive data cleaning, modeling of
heuristics like rounding and anchoring, and elimination of nonsensical
responses that might be necessary to fit a statistical model. While
statistical models are undoubtedly beneficial in many situations, we
have often found that graphical displays of experimental results are at
least as useful for analyzing and presenting the results of graphical
experiments.

@@

\caption{\label{fig-density-rug}Density of participant estimates for the
year in which the population reaches 4000. Colors are associated to
scale - linear (blue) and log (orange) - and vertical lines indicate the
true value based on the underlying model equation (black solid) and
closest point value based on the simulated data set (black dashed). A
jittered rug plot along the \(x\)-axis shows where participant estimates
were made. The plot shows anchoring occurred to the closest point as
shown by an increase in density around the dashed line. Density peaks
occurred at whole values indicating rounding errors.}

@@

If participants are making predictions and/or fitting visual statistics,
we have had success analyzing these responses by comparing the responses
to results from a statistical model to determine how visual statistics
differ from the numerical quantities derived mathematically. For
instance, we calculate the deviation between participant responses and
the linear regression in `You Draw It' experiments, then fitted
generalized additive mixed models to summarize the results across
different experimental conditions to assess how user-drawn predictions
deviated from the statistical estimates. In other direct interactions,
it may be useful to compare participant selections or annotations to
closest points on the chart to assess anchoring behavior; for discrete
selections, methods discussed in numerical estimation may also be
useful.

@@

>>>>

\title{A Guide to Designing Experiments to Test Statistical Graphics}
\author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
\date{2025-03-26}

@@

While there are heuristics, general guidelines, and best practices
(Allen \& Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on
Standards for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating
useful and visually attractive data displays, the best way to establish
the efficacy of various design decisions is to test the visualization on
humans, evaluating different variants under controlled conditions
(Cleveland et al., 1988; Cleveland \& McGill, 1985). Empirical
assessments of visualizations, when carefully designed, allow
statisticians to determine which representation of the same data is most
effective along one or more dimension(s) of interest: estimation or
prediction accuracy, within or between group comparisons, response time,
and ability to draw real-world decisions are common goals for charts.

@@

There are many different testing methods used to empirically assess
statistical graphics. This paper uses studies conducted online without
additional equipment as primary examples, though many of the same
considerations apply to in-person experiments conducted using additional
equipment, including 3D printed charts, eye-tracking equipment, and
interactive data displays. Online experiments have lower overhead, offer
relatively fast data collection, and provide useful results for
well-designed experiments. The toolkit used for these experiments is
R-based (R Core Team, 2022), and includes ggplot2 (Wickham, 2016) and
Shiny (Chang et al., 2021) as primary components. In many experiments,
we customized the Shiny interface with JavaScript and D3 (Bostock et
al., 2011), enabling interactive graphics, use of svgs, and other useful
extensions. While we prefer this set of tools, most of the observations
described here apply to a wide variety of different workflows for
graphical experimentation, including in-person experiments.

@@

Visual inference (Buja et al., 2009; Wickham et al., 2010) is another
useful testing tool for perceptual questions such as ``which chart
displays this data more clearly'' (Hofmann et al., 2012) while
simultaneously assessing the statistical significance of the graphical
finding in a chart. Visual inference charts are often called ``lineups''
in analogy to the criminal procedure where the suspect is placed in a
line with several other individuals with similar characteristics. In a
graphical lineup procedure, there is a target plot containing the real
data, embedded in an array of (typically) 19 innocent ``null'' plots
generated through resampling or simulation, for a total of 20 panels. If
viewers consistently pick the target plot at a higher rate than any of
the null plots, the target plot is said to be visually significant (Loy
\& Hofmann, 2013; Majumder et al., 2013) and a ``see'' value, the visual
analogue of a \(p\)-value (Chowdhury et al., 2020), can be calculated
using the \texttt{vinference} R package or the process described in
Vanderplas et al. (2021). The details of this calculation are beyond the
scope of this broader discussion of how to test charts, but more detail
on visual inference is provided in \textless insert citation to visual
inference WIRE article under development\textgreater.

@@

Another useful measurement strategy is to require participants to
\emph{engage directly} with an interactive visualization. This is useful
in a directed task, where users are asked to interact with the chart in
a specific way and the result is recorded, but it is also possible to
use interactive visualizations in an open-ended task, recording how
users engage with the graphic in an exploratory (as opposed to
goal-directed) manner. In one recent experiment, we asked participants
to forecast an exponential trend, with data presented on either a linear
or log scale. Using JavaScript code modified from New York Times
interactive graphics ``You Draw It'' features (Katz, 2017), we had users
draw trend lines with their computer mouse and make forecasts directly
on interactive charts, with the data and user-drawn predictions recorded
to our database (Robinson et al., 2023b). With interactive graphics
rendered using JavaScript (or other web libraries), the only limit to
the types of questions one can ask in testing graphics is one's ability
to write code to interact with the visualization library. This type of
testing method can be extremely natural for participants, but it also is
hard to generalize when discussing testing methods because of the
potential range of applications where it might be employed.

@@

All of the testing methods described here require significant work to
develop a strategy for data generation appropriate for testing the
underlying question. For instance, when testing the perception of
exponential growth, we had to develop a model which would generate data
with varying growth rates, but where the data had a pre-specified domain
and range. If the null plots fail to capture the key visual
characteristics - such as trend, spread, or clustering - then any
standout visual differences may be attributed to those unintended
features, rather than the perceptual cue being tested. In other words,
if the nulls are too obviously different, participants might detect the
real plot for the wrong reason. Each testing method has specific
requirements, but it is important to carefully calibrate the model
parameters to allow for some variability, but not too much, and to
ensure that participants can succeed at the task and do not feel like
they are being made to analyze random noise. This Goldilocks-style
problem is the focus of the next section.

@@

The choice of the parameter space used in testing is crucial to gain
insight from a study without putting too much burden on participants
with overlong studies. Choosing an appropriate space for testing
parameters is a well-known problem in psychometric testing: the space
considered should cover the area between `only some activation' to
`almost full activation' of an appropriate psychometric function (Schütt
et al., 2016; Valentin et al., 2024). When testing charts, visual
assessment is obviously key, but researchers can make use of statistical
indices related to the testing condition to narrow the parameter space
to a reasonable and efficient subset from which maximal information can
be acquired.\\
These statistical indices may also serve as quantitative proxies for the
difficulty of the visual task. To identify a statistical proxy for
visual difficulty that may help with narrowing the parameter space, it
can be useful to consider numerical measures used to estimate the same
types of visual information that will be assessed in the experiment. For
instance, we have used:

\begin{itemize}
\tightlist
\item
  \(R^2\) as a measure of the strength of a linear relationship,
\item
  Gini inequality as a measure of the strength of clustering, and
\item
  lack-of-fit statistics to assess the amount of curvature in an
  exponential relationship (shown in
  Figure~\ref{fig-lof-density-curves}).
\end{itemize}

@@

We cannot overstate the importance of visual assessment of your model
stimuli, preferably with fresh eyes. We highly recommend performing
several rounds of think-aloud pilot testing (e.g., focus groups) before
deploying an experiment. In support of this assessment, we offer up a
cautionary tale of our own experience: that of Vanderplas \& Hofmann
(2017), where we designed an experiment to test which plot aesthetics
promoted discovery of linear trends and/or clusters.

@@

Rice et al. (2024) recruited participants using nationally
representative panel samples and found that conclusions from fully
representative samples of the population can be very different from
volunteer samples recruited using other services, as many people are
unmotivated to engage with charts, don't know how to read charts, or
impose pre-formed conclusions onto visual displays. While using
representative panel sampling products offered by organizations such as
NORC or Gallup is more expensive than samples from Amazon MTurk or
Prolific, these results may imply that results from less representative
participant recruitment methods may not generalize to the broader
population. People who participate in studies via online platforms are
more technologically sophisticated and educated than the general
population, and this bias may significantly impact the conclusions.
However, it is also reasonable to argue that higher education might make
someone more likely to use charts and data for decision-making.
Researchers designing a study should carefully consider whether the goal
of the study is to generalize results to the adult population or to a
subset of that population who make decisions based on data.

@@

One highly useful (but not strictly essential) component of an
experiment that can be set up during the pilot testing stage is a basic
analysis script which summarizes all data collected to date visually. We
have used such scripts in the past to produce automatically updating
dashboards or web pages, allowing for real time or near-real time
monitoring of data collection efforts. This provides an easy way to
summarize completion of the experiment so that individuals can receive
credit (if using services like MTurk or Prolific), but also allows
interested participants to see individual results, which can be a factor
when recruiting on social media. As data collection online can also
happen extremely quickly (300 participants in \textless2 hours in our
most recent Prolific experiment), this can improve data monitoring,
allowing any issues to be spotted and resolved quickly. If server load
is a potential issue, it may also help to release batches of trials over
a longer period of time in order to minimize the chance of having to
make participants wait for others to complete the task before the server
can handle additional connections\footnote{This is the one major
  drawback to our preferred solution of self-hosting a Shiny server to
  handle data collection: the free version of Shiny server is limited to
  about 15 connections at any given time. Prolific has recently added
  rate-limiting functions to the experimental control platform, which
  makes controlling the number of active jobs much easier.}. Batch trial
releases can also be used to ensure that participants are recruited
across different time zones; in some studies this is beneficial, while
in others it may be more important to control the release time to target
individuals in e.g.~North America rather than Europe.

@@

The see-value approach (Chowdhury et al., 2020; Majumder et al., 2013;
Vanderplas et al., 2021) is extremely useful for single lineups, but
when a series of lineups that are part of a designed experiment are
used, generalized linear mixed models are a much simpler way to
summarize the overall effect of various manipulated factors (Hofmann et
al., 2012; Robinson et al., 2025; Vanderplas \& Hofmann, 2017). This
approach also works extremely well for psychometric experiments
(Vanderplas \& Hofmann, 2015), as psychometric models can be easily fit
into the framework of a generalized linear mixed model (Ju et al., 2024)
that has higher power than the Rasch models (Andrich, 1988; Lu et al.,
2022) which were historically recommended for these experiments. In
addition, similar model structures with different link functions can be
used to model accuracy, response time, and confidence, if all three
types of information are collected from participants during the
experiment.

@@

While Figure~\ref{fig-density-rug} shows point estimates, the same
approach can be modified to account for experimental methods which
generate participant response curves. In these cases, rather than adding
a rug plot to a density plot showing aggregate estimates, it may be more
helpful to create spaghetti plots of individual estimates with a
superimposed consensus estimate and relevant annotations showing
e.g.~anchor points. The advantage of this approach is that it can
accommodate extremely messy data without requiring the extensive data
cleaning, modeling of heuristics like rounding and anchoring, and
elimination of nonsensical responses that might be necessary to fit a
statistical model. While statistical models are undoubtedly beneficial
in many situations, we have often found that graphical displays of
experimental results are at least as useful for analyzing and presenting
the results of graphical experiments.

@@

\caption{\label{fig-density-rug}Density of participant estimates for the
year (x-value) in which the population reaches 4000 (y-value). Colors
are associated to scale - linear (blue) and log (orange) - and vertical
lines indicate the true value based on the underlying model equation
(black solid) and closest point value based on the simulated data set
(black dashed). A jittered rug plot along the \(x\)-axis shows where
participant estimates were made. The plot shows anchoring occurred to
the closest point as shown by an increase in density around the dashed
line. Density peaks occurred at whole values indicating rounding
errors.}

@@

If participants are making predictions and/or fitting visual statistics,
we have had success analyzing these responses by comparing the responses
to results from a statistical model to determine how visual statistics
differ from the numerical quantities derived mathematically. For
instance, in Robinson et al. (2023a), we calculate the deviation between
participant responses and the linear regression in `You Draw It'
experiments, then fitted generalized additive mixed models to summarize
the results across different experimental conditions to assess how
user-drawn predictions deviated from the statistical estimates. In other
direct interactions, it may be useful to compare participant selections
or annotations to closest points on the chart to assess anchoring
behavior; for discrete selections, methods discussed in numerical
estimation may also be useful.

@@

\bibitem[\citeproctext]{ref-robinson2023eye}
Robinson, E. A., Howard, R., \& VanderPlas, S. (2023a). Eye fitting
straight lines in the modern era. \emph{Journal of Computational and
Graphical Statistics}, \emph{32}(4), 1537--1544.

\bibitem[\citeproctext]{ref-robinson2023you}
Robinson, E. A., Howard, R., \& VanderPlas, S. (2023b). {``You draw
it''}: Implementation of visually fitted trends with r2d3. \emph{Journal
of Data Science}, \emph{21}(2), 281--294.
